Conversations in time: interactive visualisation to explore structured temporal data

### 2020-09-24 tsibbletalk.Rmd
typos

---
title: "Conversations in time: interactive visualisation to explore structured temporal data"
author:
- name: Earo Wang
  affiliation: The University of Auckland
  address: Department of Statistics
  email: earo.wang@auckland.ac.nz
- name: Dianne Cook
  affiliation: Monash University
  address: Department of Econometrics and Business Statistics
  email: dicook@monash.edu
bibliography: tsibbletalk.bib
abstract: >
  Temporal data often has a hierarchical structure, defined by categorical variables describing different levels, such as political regions or sales products. Nesting of categorical variables produces a hierarchical structure. The \pkg{tsibbletalk} package is developed to allow a user to interactively explore temporal data, relative to the nested or crossed structures. It can help to discover differences between category levels, and uncover interesting periodic or aperiodic slices. The package implements a shared `tsibble` object that allows for
  linked brushing between coordinated views, and a shiny module that aids in wrapping
  time lines for seasonal patterns. The tools are demonstrated using two data examples: retail turnover and tourism in Australia.
preamble: |
  \usepackage{subfig}
output: rticles::rjournal_article
---

@@

This article describes new software, the \CRANpkg{tsibbletalk} package, for exploring temporal data using linked views and time wrapping. The next section provides some background to the approach based on setting up data structures and workflow. The section following explains the mechanism for constructing interactivity, to link between multiple hiearchical data objects and hence plots. The section after this describes the set up for interactively slicing and dicing time to wrap a series on itself to investigate periodicities.

@@

Figure \ref{fig:highlight-retail} illustrates examining temporal data with many keys. The data has 152 series coorresponding to different items in retail data. The multiple series are displayed using an overlaid time series plot, along with a scatterplot of two variables (trend versus seasonal strength) from feature space, where each series is represented by a dot. The feature space is computed using the `features()` function from \CRANpkg{feasts}, which summarises the original data for each series using various statistical features. This function along with other **tidyverts** functions is `tsibble`-aware, and outputs a table in a reduced form where each row corresponds to a series, which can be graphically displayed as in Figure \ref{fig:highlight-retail-2}.

@@

Figure \ref{fig:highlight-retail} has also been highlighted to focus on the one series with the strongest seasonality. To create this highlighting, one needs to first filter the interesting series from the features table, and join back to the original `tsibble` in order to examine its trend in relation to others. This procedure can soon grow cumbersome if many series are to be explored. It illustrates a need to query interesting series on the fly. Although these two plots are static, we can consider them as linked views becaause the common `key` variables link between the two data tables producing the two ploots. This motivates the work in this package, described in this paper, to enable interactivity of `tsibble` and `tsibble`-derived objects for rapid exploratory data analysis.

@@

These line plots are heavily overplotted. To tease apart structure in the multiple time series, the `features()` function `fabletools` computes interesting characteristics, including the measures of trend and seasonality. These are displayed in the scatterplot at bottom right, where one dot represents one series.

@@

1. A branch of the tree corresponding to Western Australia was first selected. (The names of the regions are a little odd, which is a quirk of the data set, but all four areas, Australia's South West, ...., correspond to tourist destinations in Western Australia. Hovering over the node on the branch brings up the state name.) This generated the response in the line plots and the scatterplot that colored corresponding time series and points as blue. 
2. To enable persistent selection, in oder to compare regions or states, "Shift" and click on the tree was done, after switching the colour to red. This generated the response that points and time series corresponding to Sydney were highlighted in red. 
3. Hovering over the points brings up the label for Sydney.

@@

In summary, shared the tsibble data nicely bridges between the \CRANpkg{crosstalk} and **tidyverts** ecosystems for temporal data using the common "key". The `as_shared_tsibble()` provides a symbolic user interface for effortless construction of a hybrid of hierarchical and categorical linking between plots. The `plotly_key_tree()` fuction, in turn, decodes the hierarchical specification to plot a tree for data overview and navigation, when accompanied by more detailed plots.

@@

An important aspect of temporal data is the time context. Time has cyclical structure, that may correspond to seasonal patterns to be discovered. The `index` component of the shared tsibble data forms the basis for exploring seasonality. To investigate for periodic or aperiodic patterns, series should be wrapped on themselves, where the index is broken into temporal components like quarter or day. We'll explore this with pedestrian traffic in Melbourne, Australia.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the slider incremented by 1-day units. The "play" button near the end of slider can automately animate the slicing and dicing process, walking the viewer through all 24 hours of the 152 days. Alternatively, users can drag the slider to examine selected slices.

@@

The new shiny module exploits the temporal aspect for a tsibble object, available throu the `index` attribute. It allows users to slide through relative periods to digest seasonal behaviours, with nimble user experience.

>>>>

---
title: "Conversations in time: interactive visualisation to explore structured temporal data"
author:
- name: Earo Wang
  affiliation: The University of Auckland
  address: Department of Statistics
  email: earo.wang@auckland.ac.nz
- name: Dianne Cook
  affiliation: Monash University
  address: Department of Econometrics and Business Statistics
  email: dicook@monash.edu
bibliography: tsibbletalk.bib
abstract: >
  Temporal data often has a hierarchical structure, defined by categorical variables describing different levels, such as political regions or sales products. Nesting of categorical variables produces a hierarchical structure. The \CRANpkg{tsibbletalk} package is developed to allow a user to interactively explore temporal data, relative to the nested or crossed structures. It can help to discover differences between category levels, and uncover interesting periodic or aperiodic slices. The package implements a shared `tsibble` object that allows for linked brushing between coordinated views, and a shiny module that aids in wrapping time lines for seasonal patterns. The tools are demonstrated using two data examples: domestic tourism in Australia and foot traffic in Melbourne.
preamble: |
  \usepackage{subfig}
output: rticles::rjournal_article
---

@@

This article describes new software, the \CRANpkg{tsibbletalk} package, for exploring temporal data using linked views and time wrapping. The next section provides some background to the approach based on setting up data structures and workflow. The section following explains the mechanism for constructing interactivity, to link between multiple hierarchical data objects and hence plots. The section after this describes the set up for interactively slicing and dicing time to wrap a series on itself to investigate periodicities.

@@

Figure \ref{fig:highlight-retail} illustrates examining temporal data with many keys. The data has 152 series corresponding to different items in retail data. The multiple series are displayed using an overlaid time series plot, along with a scatterplot of two variables (trend versus seasonal strength) from feature space, where each series is represented by a dot. The feature space is computed using the `features()` function from \CRANpkg{feasts}, which summarises the original data for each series using various statistical features. This function along with other **tidyverts** functions is `tsibble`-aware, and outputs a table in a reduced form where each row corresponds to a series, which can be graphically displayed as in Figure \ref{fig:highlight-retail-2}.

@@

Figure \ref{fig:highlight-retail} has also been highlighted to focus on the one series with the strongest seasonality. To create this highlighting, one needs to first filter the interesting series from the features table, and join back to the original `tsibble` in order to examine its trend in relation to others. This procedure can soon grow cumbersome if many series are to be explored. It illustrates a need to query interesting series on the fly. Although these two plots are static, we can consider them as linked views because the common `key` variables link between the two data tables producing the two plots. This motivates the work in this package, described in this paper, to enable interactivity of `tsibble` and `tsibble`-derived objects for rapid exploratory data analysis.

@@

These line plots are heavily overplotted. To tease apart structure in the multiple time series, the `features()` function computes interesting characteristics, including the measures of trend and seasonality. These are displayed in the scatterplot at bottom right, where one dot represents one series.

@@

1. A branch of the tree corresponding to Western Australia was first selected. (The names of the regions are a little odd, which is a quirk of the data set, but all four areas, Australia's South West, ...., correspond to tourist destinations in Western Australia. Hovering over the node on the branch brings up the state name.) This generated the response in the line plots and the scatterplot that coloured corresponding time series and points as blue. 
2. To enable persistent selection, in oder to compare regions or states, "Shift" and click on the tree was done, after switching the colour to red. This generated the response that points and time series corresponding to Sydney were highlighted in red. 
3. Hovering over the points brings up the label for Sydney.

@@

In summary, shared the tsibble data nicely bridges between the \CRANpkg{crosstalk} and **tidyverts** ecosystems for temporal data using the common "key". The `as_shared_tsibble()` provides a symbolic user interface for effortless construction of a hybrid of hierarchical and categorical linking between plots. The `plotly_key_tree()` function, in turn, decodes the hierarchical specification to plot a tree for data overview and navigation, when accompanied by more detailed plots.

@@

An important aspect of temporal data is the time context. Time has cyclical structure, that may correspond to seasonal patterns to be discovered. The `index` component of the shared tsibble data forms the basis for exploring seasonality. To investigate for periodic or aperiodic patterns, series should be wrapped on themselves, where the index is broken into temporal components like quarter or day. We shall explore this with pedestrian traffic in Melbourne, Australia.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the slider incremented by 1-day units. The "play" button near the end of slider can automatically animate the slicing and dicing process, walking the viewer through all 24 hours of the 152 days. Alternatively, users can drag the slider to examine selected slices.

@@

The new shiny module exploits the temporal aspect for a tsibble object, available through the `index` attribute. It allows users to slide through relative periods to digest seasonal behaviours, with nimble user experience.

### 2020-09-24 tsibbletalk.tex
typos

\abstract{%
Temporal data often has a hierarchical structure, defined by categorical
variables describing different levels, such as political regions or
sales products. Nesting of categorical variables produces a hierarchical
structure. The \pkg{tsibbletalk} package is developed to allow a user to
interactively explore temporal data, relative to the nested or crossed
structures. It can help to discover differences between category levels,
and uncover interesting periodic or aperiodic slices. The package
implements a shared \texttt{tsibble} object that allows for linked
brushing between coordinated views, and a shiny module that aids in
wrapping time lines for seasonal patterns. The tools are demonstrated
using two data examples: retail turnover and tourism in Australia.
}

@@

Figure \ref{fig:highlight-retail} illustrates examining temporal data
with many keys. The data has 152 series coorresponding to different
items in retail data. The multiple series are displayed using an
overlaid time series plot, along with a scatterplot of two variables
(trend versus seasonal strength) from feature space, where each series
is represented by a dot. The feature space is computed using the
\texttt{features()} function from \CRANpkg{feasts}, which summarises the
original data for each series using various statistical features. This
function along with other \textbf{tidyverts} functions is
\texttt{tsibble}-aware, and outputs a table in a reduced form where each
row corresponds to a series, which can be graphically displayed as in
Figure \ref{fig:highlight-retail-2}.

@@

Figure \ref{fig:highlight-retail} has also been highlighted to focus on
the one series with the strongest seasonality. To create this
highlighting, one needs to first filter the interesting series from the
features table, and join back to the original \texttt{tsibble} in order
to examine its trend in relation to others. This procedure can soon grow
cumbersome if many series are to be explored. It illustrates a need to
query interesting series on the fly. Although these two plots are
static, we can consider them as linked views becaause the common
\texttt{key} variables link between the two data tables producing the
two ploots. This motivates the work in this package, described in this
paper, to enable interactivity of \texttt{tsibble} and
\texttt{tsibble}-derived objects for rapid exploratory data analysis.

@@

These line plots are heavily overplotted. To tease apart structure in
the multiple time series, the \texttt{features()} function
\texttt{fabletools} computes interesting characteristics, including the
measures of trend and seasonality. These are displayed in the
scatterplot at bottom right, where one dot represents one series.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A branch of the tree corresponding to Western Australia was first
  selected. (The names of the regions are a little odd, which is a quirk
  of the data set, but all four areas, Australia's South West, \ldots.,
  correspond to tourist destinations in Western Australia. Hovering over
  the node on the branch brings up the state name.) This generated the
  response in the line plots and the scatterplot that colored
  corresponding time series and points as blue.
\item
  To enable persistent selection, in oder to compare regions or states,
  ``Shift'' and click on the tree was done, after switching the colour
  to red. This generated the response that points and time series
  corresponding to Sydney were highlighted in red.
\item
  Hovering over the points brings up the label for Sydney.
\end{enumerate}

@@

In summary, shared the tsibble data nicely bridges between the
\CRANpkg{crosstalk} and \textbf{tidyverts} ecosystems for temporal data
using the common ``key''. The \texttt{as\_shared\_tsibble()} provides a
symbolic user interface for effortless construction of a hybrid of
hierarchical and categorical linking between plots. The
\texttt{plotly\_key\_tree()} fuction, in turn, decodes the hierarchical
specification to plot a tree for data overview and navigation, when
accompanied by more detailed plots.

@@

An important aspect of temporal data is the time context. Time has
cyclical structure, that may correspond to seasonal patterns to be
discovered. The \texttt{index} component of the shared tsibble data
forms the basis for exploring seasonality. To investigate for periodic
or aperiodic patterns, series should be wrapped on themselves, where the
index is broken into temporal components like quarter or day. We'll
explore this with pedestrian traffic in Melbourne, Australia.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the
slider incremented by 1-day units. The ``play'' button near the end of
slider can automately animate the slicing and dicing process, walking
the viewer through all 24 hours of the 152 days. Alternatively, users
can drag the slider to examine selected slices.

@@

The new shiny module exploits the temporal aspect for a tsibble object,
available throu the \texttt{index} attribute. It allows users to slide
through relative periods to digest seasonal behaviours, with nimble user
experience.

>>>>

\abstract{%
Temporal data often has a hierarchical structure, defined by categorical
variables describing different levels, such as political regions or
sales products. Nesting of categorical variables produces a hierarchical
structure. The \CRANpkg{tsibbletalk} package is developed to allow a
user to interactively explore temporal data, relative to the nested or
crossed structures. It can help to discover differences between category
levels, and uncover interesting periodic or aperiodic slices. The
package implements a shared \texttt{tsibble} object that allows for
linked brushing between coordinated views, and a shiny module that aids
in wrapping time lines for seasonal patterns. The tools are demonstrated
using two data examples: domestic tourism in Australia and foot traffic
in Melbourne.
}

@@

Figure \ref{fig:highlight-retail} illustrates examining temporal data
with many keys. The data has 152 series corresponding to different items
in retail data. The multiple series are displayed using an overlaid time
series plot, along with a scatterplot of two variables (trend versus
seasonal strength) from feature space, where each series is represented
by a dot. The feature space is computed using the \texttt{features()}
function from \CRANpkg{feasts}, which summarises the original data for
each series using various statistical features. This function along with
other \textbf{tidyverts} functions is \texttt{tsibble}-aware, and
outputs a table in a reduced form where each row corresponds to a
series, which can be graphically displayed as in Figure
\ref{fig:highlight-retail-2}.

@@

Figure \ref{fig:highlight-retail} has also been highlighted to focus on
the one series with the strongest seasonality. To create this
highlighting, one needs to first filter the interesting series from the
features table, and join back to the original \texttt{tsibble} in order
to examine its trend in relation to others. This procedure can soon grow
cumbersome if many series are to be explored. It illustrates a need to
query interesting series on the fly. Although these two plots are
static, we can consider them as linked views because the common
\texttt{key} variables link between the two data tables producing the
two plots. This motivates the work in this package, described in this
paper, to enable interactivity of \texttt{tsibble} and
\texttt{tsibble}-derived objects for rapid exploratory data analysis.

@@

These line plots are heavily overplotted. To tease apart structure in
the multiple time series, the \texttt{features()} function computes
interesting characteristics, including the measures of trend and
seasonality. These are displayed in the scatterplot at bottom right,
where one dot represents one series.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A branch of the tree corresponding to Western Australia was first
  selected. (The names of the regions are a little odd, which is a quirk
  of the data set, but all four areas, Australia's South West, \ldots.,
  correspond to tourist destinations in Western Australia. Hovering over
  the node on the branch brings up the state name.) This generated the
  response in the line plots and the scatterplot that coloured
  corresponding time series and points as blue.
\item
  To enable persistent selection, in oder to compare regions or states,
  ``Shift'' and click on the tree was done, after switching the colour
  to red. This generated the response that points and time series
  corresponding to Sydney were highlighted in red.
\item
  Hovering over the points brings up the label for Sydney.
\end{enumerate}

@@

In summary, shared the tsibble data nicely bridges between the
\CRANpkg{crosstalk} and \textbf{tidyverts} ecosystems for temporal data
using the common ``key''. The \texttt{as\_shared\_tsibble()} provides a
symbolic user interface for effortless construction of a hybrid of
hierarchical and categorical linking between plots. The
\texttt{plotly\_key\_tree()} function, in turn, decodes the hierarchical
specification to plot a tree for data overview and navigation, when
accompanied by more detailed plots.

@@

An important aspect of temporal data is the time context. Time has
cyclical structure, that may correspond to seasonal patterns to be
discovered. The \texttt{index} component of the shared tsibble data
forms the basis for exploring seasonality. To investigate for periodic
or aperiodic patterns, series should be wrapped on themselves, where the
index is broken into temporal components like quarter or day. We shall
explore this with pedestrian traffic in Melbourne, Australia.

@@

Figure \ref{fig:wrap-ped-1} corresponds to the initial state, with the
slider incremented by 1-day units. The ``play'' button near the end of
slider can automatically animate the slicing and dicing process, walking
the viewer through all 24 hours of the 152 days. Alternatively, users
can drag the slider to examine selected slices.

@@

The new shiny module exploits the temporal aspect for a tsibble object,
available through the \texttt{index} attribute. It allows users to slide
through relative periods to digest seasonal behaviours, with nimble user
experience.
